---
title: "Online Retail Sales Dashboard – `r params$country`"
author: "Ella Yazdi"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    css: style.css
params:
  country: All
  min_date: !r as.Date("2010-12-01")
  max_date: !r as.Date("2011-12-09")
  top_n: 10
---
# 1. Introduction

This dashboard explores customer purchasing behavior using the **Online Retail Dataset**, which contains over 500,000 transactions from a UK-based online retailer between **December 2010 and December 2011**.  

The analysis focuses on customers from **`r params$country`**, covering the time range from **`r params$min_date`** to **`r params$max_date`**.

We aim to uncover key insights such as:

- 💰 Total revenue and monthly sales trends  
- 🎯 Customer repeat behavior and segmentation  
- 📦 Best-selling and most returned products  
- 🌍 Country-level sales and return patterns  

# 2. Data Preparation (Load Packages, libraries and Raw Dataset)

**Note**: Before knitting, ensure the following packages are installed:

```{r setup, include=FALSE}
# install.packages(c("readxl", "dplyr", "lubridate", "stringr", "patchwork"))
knitr::opts_chunk$set(fig.align="center", echo = TRUE)

```


```{r load-libraries}

library(readxl)
library(dplyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(knitr)
library(scales)
library(patchwork)  # for combining plots
library(tidyr)

```


```{r load-data, message=FALSE, warning=FALSE}


retail_raw <- read_excel("Online_Retail.xlsx")

```


```{r preview-data}

glimpse(retail_raw)

```


# 3. Data Cleaning and Feature Engineering

```{r}

retail <- retail_raw %>%
  # Filter out bad rows
  filter(!is.na(CustomerID), UnitPrice > 0) %>%

  # Convert data types
  mutate(
    InvoiceNo = as.character(InvoiceNo),
    StockCode = as.character(StockCode),
    CustomerID = as.character(CustomerID),
    
    # Clean text fields
    Description = str_trim(str_to_title(Description)),

    # Create engineered features
    transaction_type = ifelse(Quantity < 0 | str_detect(InvoiceNo, "^C"), "Return", "Sale"),
    is_return = transaction_type == "Return",
    totalprice = Quantity * UnitPrice,

    # Extract date components
    invoice_date = as.Date(InvoiceDate),
    invoicemonth = floor_date(InvoiceDate, "month"),
    year_month = format(InvoiceDate, "%Y-%m"),
    weekday = factor(wday(InvoiceDate, label = TRUE), 
                     levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"), 
                     ordered = TRUE),
    hour = hour(InvoiceDate)
  )

# Clean column names for consistency
names(retail) <- names(retail) %>%
  str_replace_all("\\s+", "_") %>%
  tolower()

```

### 📌 Scope of Report

**Time Frame**: `r format(params$min_date, "%B %d, %Y")` to `r format(params$max_date, "%B %d, %Y")`  
**Country**: `r if (params$country == "All") "All Countries" else params$country`

```{r scope}
# Filter by date range
retail <- retail %>%
  filter(invoice_date >= params$min_date & invoice_date <= params$max_date)

# Apply optional country filter
retail_filtered <- if (params$country == "All") {
  retail
} else {
  retail %>% filter(country == params$country)
}

# Separate sales and returns
retail_sales <- retail_filtered %>% filter(!is_return)  # Filter only sales transactions
retail_returns <- retail_filtered %>% filter(is_return)
```

# 4. Key Metrics 

```{r Key Summary Metrics}
# Filter only sales transactions
retail_sales <- retail %>%
  filter(!is_return)

# Total Revenue
total_revenue <- sum(retail_sales$totalprice, na.rm = TRUE)

# Total Orders
total_orders <- retail_sales %>% distinct(invoiceno) %>% nrow()

# Average Order Value
aov <- total_revenue / total_orders

# Active Customers
active_customers <- n_distinct(retail_sales$customerid)

# Total Products Sold
products_sold <- sum(retail_sales$quantity)

# Repeat Purchase Rate
repeat_customers <- retail_sales %>%
  group_by(customerid) %>%
  summarise(order_count = n_distinct(invoiceno)) %>%
  filter(order_count > 1) %>%
  nrow()

repeat_rate <- repeat_customers / active_customers

# Return Rate 
return_rate <- mean(retail_filtered$is_return)
```


```{r}

# Already done:
# total_revenue, total_orders, aov, active_customers, repeat_rate, etc.

kable(data.frame(
  Metric = c("Total Revenue (£)", "Total Orders", "Average Order Value (£)", "Total Products Sold", "Active Customers", "Repeat Purchase Rate", "Return Rate"),
    Value = c(
    scales::comma(total_revenue),
    total_orders,
    round(aov, 2),
    active_customers,
    products_sold,
    scales::percent(repeat_rate, accuracy = 0.1),
    scales::percent(return_rate, accuracy = 0.1)

  )
), caption = "Overall Metrics (Entire Dataset)")

```

## Visual Insights

### 1. Time Series Trends

```{r Monthly Trends of Key Metrics, fig.width=12, fig.height=7, message = FALSE}

monthly_metrics <- retail_filtered %>%
  filter(!is.na(customerid)) %>%
  group_by(invoicemonth) %>%
  summarise(
    total_revenue = sum(totalprice, na.rm = TRUE),
    total_orders = n_distinct(invoiceno),
    active_customers = n_distinct(customerid),
    avg_order_value = total_revenue / total_orders,
    avg_orders_per_customer = total_orders / active_customers,
    .groups = "drop"
  )


# Choose metrics to compare
metrics_long <- monthly_metrics %>%
  pivot_longer(
    cols = c(total_revenue, avg_order_value, avg_orders_per_customer, active_customers),
    names_to = "Metric",
    values_to = "scaled_value"
  )

# Faceted plot
ggplot(metrics_long, aes(x = invoicemonth, y = scaled_value)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~Metric, scales = "free_y") +
  labs(
    title = "Monthly Metrics Over Time",
    x = "Month",
    y = "Scaled Value (0–1)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


```{r Monthly Return Rate}

monthly_returns <- retail %>%
  group_by(invoicemonth) %>%
  summarise(
    total_transactions = n_distinct(invoiceno),
    return_transactions = n_distinct(invoiceno[is_return]),
    return_rate = return_transactions / total_transactions,
    return_quantity = sum(abs(quantity[is_return]), na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(monthly_returns, aes(x = invoicemonth, y = return_rate)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "firebrick") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Monthly Return Rate (%) Over Time",
    x = "Month",
    y = "Return Rate (%)"
  ) +
  theme_minimal()

```


```{r Monthly Returned Quantity}

ggplot(monthly_returns, aes(x = invoicemonth, y = return_quantity)) +
  geom_line(color = "#AF7AC5", size = 1.2) +
  geom_point(color = "black", size = 2) +
  labs(title = "Monthly Returned Units",
       x = "Month",
       y = "Total Quantity Returned") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### 2. Geographical Insights
```{r Country-level revenue}

country_sales <- retail %>%
  group_by(country) %>%
  summarise(revenue = sum(totalprice, na.rm = TRUE)) %>%
  arrange(desc(revenue)) %>%
  slice_head(n = params$top_n)


ggplot(country_sales, aes(x = reorder(country, revenue), y = revenue)) +
  geom_bar(stat = "identity", fill = "tomato") +
  coord_flip() +  # flip coordinates to make it horizontal
  labs(
    title = paste("Top", params$top_n, "Countries by Revenue"),
    x = "Country",
    y = "Revenue (£)"
  ) +
  scale_y_continuous(labels = scales::label_comma()) +  # nicely format large numbers
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # center title
    axis.text = element_text(size = 10)
  )
```


### 3. Product Insights
```{r Product Performance}

# Top-selling products by quantity sold (exclude returns)
top_sales <- retail %>%
  filter(!is_return) %>%
  group_by(description) %>%
  summarise(total_sold = sum(quantity, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_sold)) %>%
  slice_head(n = params$top_n)

# Most returned products by quantity returned
top_returns <- retail %>%
  filter(is_return) %>%
  group_by(description) %>%
  summarise(total_returned = abs(sum(quantity)), .groups = "drop") %>%
  arrange(desc(total_returned)) %>%
  slice_head(n = params$top_n)


# Plot top-selling products
plot_sales <- ggplot(top_sales, aes(x = reorder(description, total_sold), y = total_sold)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = total_sold), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = paste("Top", params$top_n, "Selling Products"),
       x = "Product",
       y = "Quantity Sold") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Plot most returned products
plot_returns <- ggplot(top_returns, aes(x = reorder(description, total_returned), y = total_returned)) +
  geom_bar(stat = "identity", fill = "tomato") +
  geom_text(aes(label = total_returned), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = paste("Top", params$top_n, "Returned Products"),
       x = "Product",
       y = "Quantity Returned") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Combine plots side by side
plot_sales + plot_returns

```

### 4. Customer Segmentation (RFM)

```{r RFM Calculation and Segmentation}

rfm_data <- retail %>%
  filter(!is_return) %>%
  group_by(customerid) %>%
  summarise(
    country = first(params$country),
    recency = as.numeric(as.Date("2011-12-09") - max(invoicedate, na.rm = TRUE)),
    frequency = n_distinct(invoiceno),
    monetary = sum(totalprice, na.rm = TRUE)
  ) %>%
  ungroup()

rfm_data <- rfm_data %>%
  mutate(
    r_score = ntile(-recency, 5),
    f_score = ntile(frequency, 5),
    m_score = ntile(monetary, 5),
    rfm_score = paste0(r_score, f_score, m_score)
  ) %>%
  mutate(
    segment = case_when(
      r_score >= 4 & f_score >= 4 & m_score >= 4 ~ "Champions",
      r_score >= 4 & f_score >= 3 ~ "Loyal Customers",
      r_score >= 4 ~ "Potential Loyalists",
      r_score == 3 ~ "Recent Customers",
      r_score == 2 ~ "Customers Needing Attention",
      r_score <= 2 & f_score >= 4 ~ "At Risk",
      r_score <= 2 & f_score >= 2 ~ "Hibernating",
      TRUE ~ "Others"
    )
  )
```


```{r Distribution of Recency, Frequency, and Monetary}
library(patchwork)

p_recency <- ggplot(rfm_data, aes(x = recency)) +
  geom_histogram(fill = "skyblue", bins = 30) +
  labs(title = "Recency Distribution",
       x = "Days Since Last Purchase",
       y = "Number of Customers") +
  theme_minimal()

p_frequency <- ggplot(rfm_data, aes(x = frequency)) +
  geom_histogram(fill = "orange", bins = 30) +
  labs(title = "Frequency Distribution",
       x = "Number of Purchases",
       y = "Number of Customers") +
  theme_minimal()

p_monetary <- ggplot(rfm_data, aes(x = monetary)) +
  geom_histogram(fill = "seagreen", bins = 30) +
  labs(title = "Monetary Distribution",
       x = "Total Purchase Value (£)",
       y = "Number of Customers") +
  theme_minimal()

p_recency / p_frequency / p_monetary

```


```{r Distribution of Customers by Segment}

rfm_data %>%
  count(segment) %>%
  ggplot(aes(x = reorder(segment, -n), y = n, fill = segment)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.5) +
  labs(
    title = "Distribution of Customers by RFM Segment",
    x = "Segment",
    y = "Number of Customers"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set2")
```


```{r Return Value by RFM Segment}

retail_with_segment <- retail %>%
  filter(is_return) %>%
  inner_join(rfm_data, by = "customerid")

return_value_by_segment <- retail_with_segment %>%
  group_by(segment) %>%
  summarise(
    return_value = sum(abs(totalprice), na.rm = TRUE),
    return_count = n(),
    returned_qty = sum(abs(quantity))
  ) %>%
  arrange(desc(return_value))

ggplot(return_value_by_segment, aes(x = reorder(segment, return_value), y = return_value)) +
  geom_bar(stat = "identity", fill = "darkred") +
  geom_text(aes(label = round(return_value, 2)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(title = "Monetary Value of Returns by RFM Segment",
       x = "Customer Segment",
       y = "Return Value (£)") +
  theme_minimal()
```
